---
title: "Data SNIS"
author: "Beatriz Couto Ribeiro"
date: "2025-01-09"
output: html_document
editor_options: 
  chunk_output_type: console
---

#Limpar o ambiente 
```{r}

rm(list=ls(all=T))

```

# 1. Download Data
## 1.1. Set Folder
```{r}

directory <- "C:/Users/wb618493/OneDrive - WBG/Documents/Piaui - DPL/Cobranca de Agua/water_charges_pi"

```

## 1.2. Download Packages
```{r setup, include=FALSE}
suppressMessages(memory.limit(size = NA))

if(!require(install.load)){
  install.packages("install.load")
  library(install.load)
}
suppressMessages(install_load("tidyverse", "dplyr", "ggplot2", "readr", "readxl","knitr","here", "data.table","basedosdados","bigrquery", "dbplyr","janitor","fuzzyjoin","zoo", "purrr", "stringr","DBI", "openxlsx"))

`%notin%` <- Negate(`%in%`)

```

## 1.3. set the google cloud console ID and project
```{r}

set_billing_id("snis-data")

```

# 2. Download Dataset
```{r}

#snis_all <- readRDS("snis_br_add.RDS")

```


# 2.1. Connect to RAIS database
```{r}

# Para carregar o dado direto no R
query <- "uf"
uf_pop_data <- basedosdados::bdplyr(query)

# Check data structure
#dplyr::glimpse(snis_municipio_data)

```

## 2.2. Extract data from "Serviços de Água e Esgoto nos Municípios": https://basedosdados.org/dataset/2a543ad8-3cdb-4047-9498-efe7fb8ed697?table=df7cf198-4889-4baf-bb77-4e0e28eb90ca
```{r}

# Read the expanded query into a dataframe
# snis_all <- basedosdados::read_sql(query = "SELECT
#     dados.ano as ano,
#     dados.sigla_uf AS sigla_uf,
#     diretorio_sigla_uf.nome AS sigla_uf_nome,
#     dados.populacao_atendida_agua as populacao_atendida_agua,
#     dados.populacao_atentida_esgoto as populacao_atentida_esgoto,
#     dados.populacao_urbana as populacao_urbana,
#     dados.populacao_urbana_residente_agua as populacao_urbana_residente_agua,
#     dados.populacao_urbana_atendida_agua as populacao_urbana_atendida_agua,
#     dados.populacao_urbana_residente_esgoto as populacao_urbana_residente_esgoto,
#     dados.populacao_urbana_atendida_esgoto as populacao_urbana_atendida_esgoto,
#     dados.volume_esgoto_coletado as volume_esgoto_coletado,
#     dados.volume_esgoto_tratado as volume_esgoto_tratado,
#     dados.volume_esgoto_faturado as volume_esgoto_faturado,
#     dados.indice_perda_faturamento as indice_perda_faturamento,
#     dados.indice_perda_distribuicao_agua as indice_perda_distribuicao_agua
# FROM `basedosdados.br_mdr_snis.municipio_agua_esgoto` AS dados
# LEFT JOIN (SELECT DISTINCT sigla,nome  FROM `basedosdados.br_bd_diretorios_brasil.uf`) AS diretorio_sigla_uf
#     ON dados.sigla_uf = diretorio_sigla_uf.sigla
# ")

#Read the expanded query into a dataframe
snis_all_add <- basedosdados::read_sql(query = "SELECT
    dados.ano as ano,
    dados.sigla_uf AS sigla_uf,
    diretorio_sigla_uf.nome AS sigla_uf_nome,
    dados.populacao_atendida_agua as populacao_atendida_agua,
    dados.populacao_atentida_esgoto as populacao_atentida_esgoto,
    dados.populacao_urbana as populacao_urbana,
    dados.populacao_urbana_residente_agua as populacao_urbana_residente_agua,
    dados.populacao_urbana_atendida_agua as populacao_urbana_atendida_agua,
    dados.populacao_urbana_residente_esgoto as populacao_urbana_residente_esgoto,
    dados.populacao_urbana_atendida_esgoto as populacao_urbana_atendida_esgoto,
    dados.volume_esgoto_coletado as volume_esgoto_coletado,
    dados.volume_esgoto_tratado as volume_esgoto_tratado,
    dados.volume_esgoto_faturado as volume_esgoto_faturado,
    dados.indice_perda_faturamento as indice_perda_faturamento,
    dados.indice_coleta_esgoto as indice_coleta_esgoto,
    dados.indice_tratamento_esgoto as indice_tratamento_esgoto,
    dados.indice_atendimento_urbano_agua as indice_atendimento_urbano_agua,
    dados.indice_atendimento_esgoto_esgoto as indice_atendimento_esgoto_esgoto,
    dados.indice_perda_distribuicao_agua as indice_perda_distribuicao_agua,
    dados.indice_atendimento_total_agua as indice_atendimento_total_agua
FROM `basedosdados.br_mdr_snis.municipio_agua_esgoto` AS dados
LEFT JOIN (SELECT DISTINCT sigla,nome  FROM `basedosdados.br_bd_diretorios_brasil.uf`) AS diretorio_sigla_uf
    ON dados.sigla_uf = diretorio_sigla_uf.sigla
")

#Save database
#saveRDS(snis_all_add,"snis_br_add.RDS")

```


## 2.3. Extract data from "População Brasileira": https://basedosdados.org/dataset/d30222ad-7a5c-4778-a1ec-f0785371d1ca?table=b99f0017-e587-477e-8cfb-05fb5d1005b8
```{r}

# Para carregar o dado direto no R
# pop_data <- basedosdados::read_sql(query = "SELECT
#     dados.sigla_uf AS sigla_uf,
#     diretorio_sigla_uf.nome AS sigla_uf_nome,
#     dados.ano as ano,
#     dados.populacao as populacao
# FROM `basedosdados.br_ibge_populacao.uf` AS dados
# LEFT JOIN (SELECT DISTINCT sigla,nome  FROM `basedosdados.br_bd_diretorios_brasil.uf`) AS diretorio_sigla_uf
#     ON dados.sigla_uf = diretorio_sigla_uf.sigla
# ")

#Save database
#saveRDS(pop_data,"pop_data.RDS")


```


# 3. Data Analysis - PI
```{r}

# Select only PI
snis_pi <- subset(snis_all, sigla_uf == "PI")

pop_data_pi <- subset(pop_data, sigla_uf == "PI")

# Select some columns 
snis_pi <- snis_pi %>% select(1:2, 4:20)

# Summarise
snis_pi_summary <- snis_pi %>%
  group_by(ano, sigla_uf) %>%
  summarise(
    # Summing the specified variables
    populacao_atendida_agua = sum(populacao_atendida_agua, na.rm = TRUE),
    populacao_atentida_esgoto = sum(populacao_atentida_esgoto, na.rm = TRUE),
    populacao_urbana = sum(populacao_urbana, na.rm = TRUE),
    populacao_urbana_residente_agua = sum(populacao_urbana_residente_agua, na.rm = TRUE),
    populacao_urbana_atendida_agua = sum(populacao_urbana_atendida_agua, na.rm = TRUE),
    populacao_urbana_residente_esgoto = sum(populacao_urbana_residente_esgoto, na.rm = TRUE),
    populacao_urbana_atendida_esgoto = sum(populacao_urbana_atendida_esgoto, na.rm = TRUE),
    volume_esgoto_coletado = sum(volume_esgoto_coletado, na.rm = TRUE),
    volume_esgoto_tratado = sum(volume_esgoto_tratado, na.rm = TRUE),
    volume_esgoto_faturado = sum(volume_esgoto_faturado, na.rm = TRUE),
    
    # Averaging the specified variables
    indice_perda_faturamento = mean(indice_perda_faturamento, na.rm = TRUE),
    indice_coleta_esgoto = mean(indice_coleta_esgoto, na.rm = TRUE),
    indice_tratamento_esgoto = mean(indice_tratamento_esgoto, na.rm = TRUE),
    indice_atendimento_urbano_agua = mean(indice_atendimento_urbano_agua, na.rm = TRUE),
    indice_atendimento_esgoto_esgoto = mean(indice_atendimento_esgoto_esgoto, na.rm = TRUE),
    indice_atendimento_total_agua = mean(indice_atendimento_total_agua, na.rm = TRUE),
    indice_perda_distribuicao_agua = mean(indice_perda_distribuicao_agua, na.rm = TRUE)
  )


snis_pi_pop_summary <- snis_pi_summary %>% left_join(pop_data_pi %>% select(sigla_uf, ano, populacao), by = c("sigla_uf", "ano"))


write_csv(snis_pi_pop_summary,"snis_pi.csv")

```


# 4.Plots
# Total number of TOTAL population with water and saniation services in PI
```{r}

# Reshape the data to long format for ggplot
snis_pi_long <- snis_pi_summary %>%
  pivot_longer(cols = c(populacao_atendida_agua, populacao_atentida_esgoto), 
               names_to = "variable", 
               values_to = "value")

# Create the line graph
ggplot(snis_pi_long, aes(x = ano, y = value, color = variable, group = variable)) +
  geom_line(size = 1) +  # Create the lines
  geom_point(size = 2) + # Optionally add points for clarity
  labs(
    title = "Population Attended by Water and Sewage Services",
    x = "Year (Ano)",
    y = "Population Attended",
    color = "Service Type"
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 3),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(2005, 2022)             # Ensure the axis spans from 2005 to 2023
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",  # Position the legend at the top
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )


```


# Total number of URBAN population with water and saniation services in PI
```{r}

# Reshape the data to long format for ggplot
snis_pi_long <- snis_pi_summary %>%
  pivot_longer(cols = c(populacao_urbana_atendida_agua, populacao_urbana_atendida_esgoto), 
               names_to = "variable", 
               values_to = "value")

# Create the line graph
ggplot(snis_pi_long, aes(x = ano, y = value, color = variable, group = variable)) +
  geom_line(size = 1) +  # Create the lines
  geom_point(size = 2) + # Optionally add points for clarity
  labs(
    title = "Population Attended by Water and Sewage Services",
    x = "Year (Ano)",
    y = "Population Attended",
    color = "Service Type"
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 3),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(2005, 2022)             # Ensure the axis spans from 2005 to 2023
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",  # Position the legend at the top
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )


```

# Indice de coleta de esgoto e atendimento agua   
```{r}

# Reshape the data to long format for ggplot
snis_pi_long <- snis_pi_summary %>%
  pivot_longer(cols = c(indice_coleta_esgoto, indice_atendimento_total_agua), 
               names_to = "variable", 
               values_to = "value")

# Create the line graph
ggplot(snis_pi_long, aes(x = ano, y = value, color = variable, group = variable)) +
  geom_line(size = 1) +  # Create the lines
  geom_point(size = 2) + # Optionally add points for clarity
  geom_text(aes(label = paste0(round(value, 1), "%")), vjust = -0.5, size = 3) +  # Add values as percentage on top of the points
  labs(
    title = "Population Attended by Water and Sewage Services",
    x = "Year (Ano)",
    y = "Population Attended",
    color = "Service Type"
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 10),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(0, 100)             # Ensure the axis spans from 2005 to 2023
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 3),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(2005, 2022)             # Ensure the axis spans from 2005 to 2023
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",  # Position the legend at the top
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )

```



# Indice de tratamento de esgoto  
```{r}

# Reshape the data to long format for ggplot
snis_pi_long <- snis_pi_summary %>%
  pivot_longer(cols = c(indice_tratamento_esgoto), 
               names_to = "variable", 
               values_to = "value")

# Create the line graph
ggplot(snis_pi_long, aes(x = ano, y = value, color = variable, group = variable)) +
  geom_line(size = 1) +  # Create the lines
  geom_point(size = 2) + # Optionally add points for clarity
    geom_text(aes(label = paste0(round(value, 1), "%")), vjust = -0.5, size = 3) +  # Add values as percentage on top of the points
  labs(
    title = "Population Attended by Water and Sewage Services",
    x = "Year (Ano)",
    y = "Population Attended",
    color = "Service Type"
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 10),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(0, 100)             # Ensure the axis spans from 2005 to 2023
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 3),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(2005, 2022)             # Ensure the axis spans from 2005 to 2023
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",  # Position the legend at the top
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )

```


# Perda distribuicao 
```{r}

# Reshape the data to long format for ggplot
snis_pi_long <- snis_pi_summary %>%
  pivot_longer(cols = c(indice_perda_distribuicao_agua), 
               names_to = "variable", 
               values_to = "value")

# Create the line graph
ggplot(snis_pi_long, aes(x = ano, y = value, color = variable, group = variable)) +
  geom_line(size = 1) +  # Create the lines
  geom_point(size = 2) + # Optionally add points for clarity
      geom_text(aes(label = paste0(round(value, 1), "%")), vjust = -0.5, size = 3) +  # Add values as percentage on top of the points
  labs(
    title = "Population Attended by Water and Sewage Services",
    x = "Year (Ano)",
    y = "Population Attended",
    color = "Service Type"
  ) +
  scale_y_continuous(
    breaks = seq(0, 100, by = 10),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(0, 100)             # Ensure the axis spans from 2005 to 2023
  ) +
  scale_x_continuous(
    breaks = seq(2005, 2022, by = 3),  # Set x-axis breaks from 2005 to 2023 with 3-year intervals
    limits = c(2005, 2022)             # Ensure the axis spans from 2005 to 2023
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",  # Position the legend at the top
    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate x-axis labels for readability
  )

```



